\texorpdfstring{\ZpL}{ZpL}: a p-adic precision package

### 2018-05-04 ZpL/ZpL.tex
Address some reviewers' remarks

\newcommand{\sage}{\textsc{SageMath}\xspace}
\newcommand{\ZpCR}{\text{\color{output} \rm \tt ZpCR}\xspace}
\newcommand{\ZpFP}{\text{\color{output} \rm \tt ZpFP}\xspace}
\newcommand{\ZpL}{\text{\color{output} \rm \tt ZpL}\xspace}
\newcommand{\ZpLC}{\text{\color{output} \rm \tt ZpLC}\xspace}
\newcommand{\ZpLF}{\text{\color{output} \rm \tt ZpLF}\xspace}
\newcommand{\ZpXX}{\text{\color{output} \rm \tt ZpXX}\xspace}
\newcommand{\QpXX}{\text{\color{output} \rm \tt QpXX}\xspace}

@@

In this paper, we describe an implementation of this idea in \sage \cite{sage}.
Rather than attaching a precision to each element, we store the precision of
many elements together by tracking a precision module for the whole collection
of variables.  As variables are created and destroyed, we update a matrix whose
rows represent the vectors in the module.  Information about the precision
of elements is extracted from the matrix as necessary.

@@

In \S \ref{sec:implementation} we give more details on the implementation.
\S \ref{ssec:preclemma} contains a brief overview on the theory of
$p$-adic precision of \cite{caruso-roe-vaccon:14a}. 
In the next two subsections, we explain in more details how \ZpLC and 
\ZpLF work. \S \ref{ssec:trackprec} is devoted to the implementation of 
automatic differentiation leading to the actual computation of the
module that models the precision.
In \S \ref{ssec:viewprec}, we explain how precision on any individual number
can be recovered and discuss the validity
of our results.
The complexity overhead induced by our package is analyzed in
\S \ref{ssec:complexity}.

@@

\noindent
\ZpXX is a generic notation for \ZpCR, \ZpLC and \ZpLF.
The first, \ZpCR, is the usual constructor for $p$-adic parents in 
\sage.\footnote{It stands for \textit{capped relative}.} It tracks precision using interval arithmetic. On the contrary
\ZpLC and \ZpLF are provided by our package. In the sequel, we will
compare the outputs provided by each parent.
As the results for \ZpLF and \ZpLC agree on all examples in this section, so we only write \ZpLC for them. 
More details about what are precisely \ZpLC and \ZpLF are provided in Section \ref{sec:implementation}.

@@

We begin our tour of the features of the \ZpL package with some basic 
arithmetic computations.
We first pick some random element $x$ (using the natural
Haar measure on $\mathbb{Z}_p$, as is default in \sage).
The function {\color{function}\verb?random_element?} is designed so
that it guarantees that the picked random element is the same for
each constructor \ZpCR, \ZpLC and \ZpLF.

@@

\noindent \textbf{Euclidean algorithm.}
%
A natural example is that of the computation of GCD, whose stability has 
been studied in \cite{caruso:2017}.
A naive implementation of the Euclidean algorithm can produce different 
behavior depending on the type of implementation of the field of 
$p$-adic coefficients.

@@

Taking the compositum of all the functions $\tilde F$ (\emph{cf} 
Eq.~\eqref{eq:tildeF}) from time $0$ to $t$, we find that the execution 
of the session until time $t$ is modeled by a mathematical function
$\Phi_t : \calU_t \oplus \calK_t \to E_t$.
From the design of \ZpLC, we deduce further that there exists a 
vector $\underline k_t \in K_t$ such that:
$$\Phi_t(\underline u_t \oplus \underline k_t) = \underline x_t 
\quad \text{and} \quad
d\Phi_t (U_t \oplus K_t) = H_t$$
where the differential of $\Phi_t$ is taken at the point
$\underline u_t \oplus \underline k_t$. 
Set $\Phi_{t,\ttv} = \pi_\ttv \circ \Phi_t$; it maps 
$\underline u_t \oplus \underline k_t$ to the \ttv-coordinate 
$x_{t,\ttv}$ of $\underline x_t$ and satisfies
$d\Phi_{t,\ttv} (U_t \oplus K_t) = \pi_\ttv(H_t) = p^N \Zp$
where $N$ is the value returned by {\color{method}\verb?precision_absolute?}. Thus, 
as soon as the assumptions of Proposition \ref{prop:precision} are
fulfilled, we derive
$\Phi_{t,\ttv} \big((\underline u_t + U_t) \oplus (\underline k_t + K_t)\big) = 
x_{t,\ttv} + p^N \Zp$.
Noting that $k_t \in K_t$, we finally get:
\begin{equation}
\label{eq:absprec}
\Phi_{t,\ttv} (\underline u_t + U_t) \subset
\Phi_{t,\ttv} \big((\underline u_t + U_t) \oplus K_t\big) = 
x_{t,\ttv} + p^N \Zp.
\end{equation}
The latter inclusion means that the computed value $x_{t,\ttv}$ is
accurate at precision $O(p^N)$, \emph{i.e.} that the output absolute
precision is correct.

>>>>

\newcommand{\sage}{\textsc{SageMath}\xspace}
\newcommand{\ZpCR}{\text{\color{output} \rm \tt ZpCR}\xspace}
\newcommand{\ZpFP}{\text{\color{output} \rm \tt ZpFP}\xspace}
\newcommand{\ZpL}{\text{\color{output} \rm \tt ZpL}\xspace}
\newcommand{\ZpLC}{\text{\color{output} \rm \tt ZpLC}\xspace}
\newcommand{\ZpLF}{\text{\color{output} \rm \tt ZpLF}\xspace}
\newcommand{\ZpXX}{\text{\color{output} \rm \tt ZpXX}\xspace}
\newcommand{\QpLC}{\text{\color{output} \rm \tt QpLC}\xspace}
\newcommand{\QpLF}{\text{\color{output} \rm \tt QpLF}\xspace}
\newcommand{\QpXX}{\text{\color{output} \rm \tt QpXX}\xspace}

\newcommand{\LC}{\text{\rm \tt LC}\xspace}
\newcommand{\LF}{\text{\rm \tt LF}\xspace}

@@

In this paper, we describe an implementation of this idea in \sage \cite{sage}.
Rather than attaching a precision to each element, we store the precision of
many elements together by tracking a precision module for the whole collection
of variables.  As variables are created and destroyed, we update a matrix whose
rows represent the vectors in the module.  Information about the precision
of elements is extracted from the matrix as necessary.
Concretely, our package provides two new \sage 
parents\footnote{Parents in \sage are the analogues of container objects 
in computer science. Basically it corresponds to the implementation of a 
concrete mathematical structure, as the ring of $p$-adic integers.}, 
namely \ZpLC and \ZpLF (and their fraction fields \QpLC and \QpLF). The 
acronyms \LC and \LF stand for ``lattice-cap'' and ``lattice-float''
respectively (see \S \ref{sec:implementation} for details). Our package 
is merged in the standard distribution of \sage since version 8.2.

@@

In \S \ref{sec:implementation} we give more details on the implementation.
\S \ref{ssec:preclemma} contains a brief overview on the theory of
$p$-adic precision of \cite{caruso-roe-vaccon:14a}. 
In the next two subsections, we explain in more details how our package
works. \S \ref{ssec:trackprec} is devoted to the implementation of 
automatic differentiation leading to the actual computation of the
module that models the precision.
In \S \ref{ssec:viewprec}, we explain how precision on any individual number
can be recovered and discuss the validity
of our results.
The complexity overhead induced by our package is analyzed in
\S \ref{ssec:complexity}.

@@

\noindent
\ZpXX is a generic notation for \ZpCR, \ZpLC and \ZpLF.
The first, \ZpCR, is the usual constructor for $p$-adic parents in 
\sage\footnote{It stands for \textit{capped relative}.}. It tracks 
precision using interval arithmetic. On the contrary
\ZpLC and \ZpLF are provided by our package. In the sequel, we will
compare the outputs provided by each parent.
As the results for \ZpLF and \ZpLC agree on all examples in this section, 
we shall only mention \ZpLC in the sequel.
More details about what are precisely \ZpLC and \ZpLF are provided in Section \ref{sec:implementation}.

In what follows, random elements in $\Zp$ are always picked with respect 
to the Haar measure. Random elements in $\Qp$ are constructed as a 
product $p^V \cdot U$ where $U$ is a random element in $\Zp$ 
(distributed according to the Haar measure) and $V$ is a random variable
taking values in $\Z$ which is independant from $U$ and distributed as
follows:
$$\textstyle
\text{Prob}[V{=}0] = \frac 1 5, \quad
\text{Prob}[V{=}v] = \frac 2 {5|v|\cdot (|v|+1)}, \, \forall v \neq 0.$$
This distribution is the default in \sage.

@@

We begin our tour of the features of the \ZpL package with some basic 
arithmetic computations.
We first pick some random element $x$.
The function {\color{function}\verb?random_element?} is designed so
that it guarantees that the picked random element is the same for
each constructor \ZpCR, \ZpLC and \ZpLF.

@@

\noindent \textbf{Euclidean algorithm.}
%
A natural example is that of the computation of GCD in the Euclidean 
ring $\Qp[X]$, whose stability has been studied in \cite{caruso:2017}. A 
naive implementation of the Euclidean algorithm can produce different 
behavior depending on the type of implementation of the field of 
$p$-adic coefficients.

@@

Taking the composition of all the functions $\tilde F$ (\emph{cf} 
Eq.~\eqref{eq:tildeF}) from time $0$ to $t$, we find that the execution 
of the session until time $t$ is modeled by a mathematical function
$\Phi_t : \calU_t \oplus \calK_t \to E_t$.
From the design of \ZpLC, we deduce further that there exists a 
vector $\underline k_t \in K_t$ such that:
$$\Phi_t(\underline u_t \oplus \underline k_t) = \underline x_t 
\quad \text{and} \quad
d\Phi_t (U_t \oplus K_t) = H_t$$
where the differential of $\Phi_t$ is taken at the point
$\underline u_t \oplus \underline k_t$. 
Set $\Phi_{t,\ttv} = \pi_\ttv \circ \Phi_t$; it maps 
$\underline u_t \oplus \underline k_t$ to the \ttv-coordinate 
$x_{t,\ttv}$ of $\underline x_t$ and satisfies
$d\Phi_{t,\ttv} (U_t \oplus K_t) = \pi_\ttv(H_t) = p^N \Zp$
where $N$ is the value returned by {\color{method}\verb?precision_absolute?}. Thus, 
as soon as the assumptions of Proposition \ref{prop:precision} are
fulfilled, we derive
$\Phi_{t,\ttv} \big((\underline u_t + U_t) \oplus (\underline k_t + K_t)\big) = 
x_{t,\ttv} + p^N \Zp$.
Noting that $k_t \in K_t$, we finally get:
\begin{equation}
\label{eq:absprec}
\Phi_{t,\ttv} (\underline u_t + U_t) \subset
\Phi_{t,\ttv} \big((\underline u_t + U_t) \oplus K_t\big) = 
x_{t,\ttv} + p^N \Zp.
\end{equation}
The latter inclusion means that the computed value $x_{t,\ttv}$ is
accurate at precision $O(p^N)$, \emph{i.e.} that the output absolute
precision is correct.
